\section{Related Work}
\label{sec:related}

Three systems in the literature focus on systematic support of
evolving graphs, all of them non-compositional.  \eat{None of the three
systems are publicly available, so a direct comparison with them is
impossible.}

Miao et al.~\cite{Miao2015} developed ImmortalGraph, a proprietary
in-memory execution engine for temporal graph analytics.
It\eat{ImmortalGraph} does not provide a query language, focusing
primarily on efficient physical data layout.  In contrast, \sys is an
open source system and supports the full TGA. \eat{ The batching
  method for snapshot analytics we use is similar to the one proposed
  in ImmortalGraph.  However,} Further, ImmortalGraph was developed
with the focus on centralized rather than distributed computation, but
we leverage their insights about data locality, adapting them for
HDFS.

The G* system~\cite{Labouseur2015} manages graphs that correspond to
periodic snapshots, with the focus on efficient data layout.  It takes
advantage of the similarity between successive snapshots by storing
shared vertices once and maintaining per-graph indexes. G* provides
two query languages: procedural PGQL and declarative DGQL. Time is not
an intrinsic part of the G* model, and thus temporal queries with time
predicates are not supported.  In contrast, \sys supports temporal
window node creation, temporal subgraph, and other intrinsically
time-based operations.  \eat{ PGQL supports graph operators like
  retrieving vertices and their edges from disk and non-graph
  operators like aggregate, union, projection, and join.}\eat{ All
  operators use a streaming model, i.e., like in a traditional DBMS,
  they pipeline.}\eat{ DGQL is similar to SQL and is converted into
  PGQL by the system.}

Finally, the Historical Graph Store (HGS) is an evolving graph query
system based on Spark.  It uses the property graph model and supports
retrieval tasks along time and entity dimensions through Java/Python
APIs.  HGS supports slice, selection and nodecompute (like \tga
subgraph and map, respectively, but with no temporal predicates), and
various evolution-centered operators.  In contrast, in addition to a
principled support for temporal predicates, \sys also supports node
and edge creation operations which allow for complex analyses such as
journeys and aggregations of desired structural and temporal
resolution.

\eat{HGS does not provide formal
semantics for its operations, the main focus is on efficient on-disk
representation for retrieval.}
