\section{Introduction}
\label{sec:intro}

% General motivation
The importance of networks in scientific and commercial domains cannot
be overstated.  Considerable research and engineering effort is being
devoted to developing effective and efficient graph representations
and analytics.  Efficient graph abstractions and analytics for {\em
  static graphs} are available to researchers and practitioners in
scope of open source platforms such as Apache Giraph, Apache Spark /
GraphX~\cite{DBLP:conf/osdi/GonzalezXDCFS14} and GraphLab /
PowerGraph~\cite{DBLP:conf/osdi/GonzalezLGBG12}.

Arguably the most interesting and important questions one can ask
about networks have to do with their evolution, rather than with their
static state.  Analysis of {\em evolving graphs} has been receiving
increasing
attention~\cite{DBLP:journals/csur/AggarwalS14,Chan2008,Kan2009,Miao2015,Ren2011,Semertzidis2015}.
Yet, despite much recent activity, and despite increased variety and
availability of evolving graph data, systematic support for scalable
querying and analysis of evolving graphs still lacks.  This support is
urgently needed, due first and foremost to the scalability and
efficiency challenges inherent in evolving graph analysis, but also to
considerations of usability and ease of dissemination.  The goal of
our work is to fill this gap.  In this paper, we present an algebraic
query language for exploratory analysis of evolving graph, and its
scalable implementation in \ql, an open-source distributed framework
built on top of Apache Spark.

Our goal in developing \tga is to give users an ability to concisely
express a wide range of common analysis tasks, while at the same time
preserving inter-operability with standard SQL.  Further, we aim to
enable an implementation of our algebraic framework on top of a
relational engine. \julia{this is still not convincing}

Implementing (non-temporal) graph querying and analytics in an RDBMS
has been receiving renewed
attention~\cite{DBLP:conf/sigmod/AbergerTOR16,DBLP:conf/sigmod/SunFSKHX15,DBLP:journals/pvldb/Xirogiannopoulos15},
and our work is in-line with this trend. Our data model is based on
the temporal relational model, and our algebra corresponds to temporal
relational algebra, but is designed specifically for evolving graphs.
We give a translation of our operators in relational algebra, and, as
consequence, can implement our algebraic framework on top of a
relational engine.  Th eadvantages of incorporating graph querying
into a general-purpose relational system are two-fold: usability and
access to the relational data modeling and query processing
infrastructure. 

For this reason our data model is based on the temporal relational
model, and our query language corresponds to temporal relational
calculus, a two-sorted logic with variables and quantifiers explicitly
ranging over the time domain~\cite{DBLP:reference/db/Toman09}, for
graphs.

Specifically, we represent graph evolution, including changes in
topology or attribute values of vertices and edges, using the \tg
abstraction --- a collection of temporal SQL relations with
appropriate integrity constraints.  An example of a \tg is given in
Figure~\ref{fig:tg_ve}, where we show evolution of a co-authorship
network.  

\tga does not support general recursion or transitive closure
computation. (Although, as we will see in Section~\ref{sec:analytics},
Pregel-style graph analytics such as PageRank are supported as an
extension.)  For this reason, we also do not support regular path
queries (RPQ) or the more general path query classes (CRPQ and ECRPQ).
Extending our formalism with recursion and path queries is
non-trivial, and we leave this to future work, see
Section~\ref{sec:conc} for a discussion.

Rather than focusing on path computation and graph traversal, we
stress tasks that perform whole-graph analysis over time.  Several
such tasks were described next, in Section~\ref{sec:cases}.
Additional examples can be found in SocialScope~\cite{Amer-Yahia2009}
--- a closed non-temporal algebra for multigraphs that is motivated by
information discovery in social content sites.  It is not difficult to
show that the graph (rather than multi-graph) versions of all
SocialScope operations can be expressed, and augmented with the
temporal dimension, in \tga.

We now present several motivating scenarios that highlight
the features of \ql algebra.

\input{usecases}

%\input{design}

\subsection{Contributions and roadmap}

We make the following contributions to support efficient and usable
analysis of evolving graphs.

\begin{enumerate}[noitemsep,leftmargin=*]
%\begin{description}[noitemsep]

\item We propose a representation of an evolving graph, called a \tg,
  which captures the evolution of both graph topology and vertex and
  edge attributes (Section~\ref{sec:model}), and develop a
  compositional \tg algebra, \tga (Section~\ref{sec:algebra}).

\item We show a reduction from \tga to temporal relational algebra
  \tra, using a combination of standard operators and \tg-specific
  primitives, and present formal properties of \tga
  (Section~\ref{sec:formal}).

\item We present an implementation in scope of the \ql system, built
  on Apache Spark / GraphX~\cite{DBLP:conf/osdi/GonzalezXDCFS14}. \ql
  supports several access methods that correspond to different
  trade-offs in temporal and structural locality
  (Section~\ref{sec:sys}).

\item We conduct an extensive experimental evaluation with real
  datasets, demonstrating that \ql scales (Section~\ref{sec:exp}).  We
  also illustrate the usability throughout the paper, with a variety
  of real-life analysis tasks that can be concisely expressed in our
  algebra.

\end{enumerate}


